draft 1

### 2016-06-20 finalwriteup/draft.rmd
almost finished method review

ANOVA - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.
B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
DAG - directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

@@

ANOVA is closely related to linear fit models but incorporates the analysis of differences in group means (i.e. categorical variables) [@Lunney1970]. The method is simple and fast but assumes variables are normally distributed and linearlly related. The variables must therefore be normalised before analysis, however it cannot be assumed that each variable can be perfectly normalised. A disadvantage of this method is that linear relationships and normal variables do not necessarily represent real world data in the best way [@Breiman2001b].

In order to compare variable importance, the output of the model is assessed which includes the linear coefficients for each variable and the p-values for the coefficients [@Markham2016]. The magnitude of the coefficient does not indicate importance relative to the other variables because the magintude is directly dependent on the values within the variable. This leaves the p-values which are the result of a test for whether there is a significant relationship between the variable and the target variable. This is a good indication of whether the variable has a relationship with the target variable, however doesnt necessarily rank importance. If a variable is measured more precisely, it will have a smaller p value, whereas a variable measured roughly will have a higher p value. This does not necessarily make the more precisely measure variable more important than the other. The p test does however distinguish between variables that have a statistical relationship to the target variable and ones that do not. With regards to p-values it is also important to keep in mind that, with a dozen variables, chance alone can produce a variable with a pvalue < 0.001 7% of the time [@Rice1989].

@@

The random forest algorithm can produce a permutation variable importance for each variable as part of its output. Variable importance is represented as a score, and to do this a permutation and calcuations are performed on each variable. More specifically, consider a single covariate. The values of this variable in the out of bag sample are randomly permuted (reordered) for each tree - a technique that mimics the absence of that variable. The trees in the random forest are then run again. Variable importance can now be derived by comparing the results of the forest where the covariate has been permuted to the original random forest. If the accuracy of the premuted forest is much less than the original forest, then that variable is important. Importance scores are calculated as the mean decrease in accuracy over all trees for that permuted variable in the random forest [@Breiman2005].

The random forest variable importance provides valuable insight because it addresses the impact of each predictor variable individually as well as in multivariate interactions with other predictor variables. It handles uncorrelated variables well if the samples for the trees are subsampled without replacement [@Strobl2007]. However, a severe disadvantage of random forest variable importance is that it is not reliable for variable selection where potential predictor variables vary in their scale of measurement or their number of categories. The importance of correlated predictors is overestimated and the algorithm tends to select variables that have many possible splits or many missing values [@Strobl2007]. Overall, the random forest permutation importance is useful as it is able to compare variables in a machine learning environment and can include multivariate interactions, but the current random forest importance measure overestimates the importance of variables with many categories and some numeric variables.

@@

Cforests are an alterantive ensemble tree method to random forests that overcome the aforementioned shortfalls of random forests such as overestimating the importance of variables with many categories and numeric variables. It is built from ctrees that are based on a conditional inference framework. The key difference between ctrees and standard decision trees are that a significance test procedure for splitting instead of an purity meausre such as the Gini coefficient, which is centered around information gain.

The details behind the significance test procedure of a ctree are as follows. Each variable is permuted in every possible way, and a correlation value is calucated between the tested variable and target variable, for each permutation. The unchanged variable correlation is then compared with the correlation values for all permutations of that variable. From this, a p-value for the true correlation value can be calculated. The predictor variable with the lowest p-value is selected as the splitting variable [@Hothorn2006]. Using an ensemble forest of these trees, permutation is then calculated in the same fashion as random forests, via permuting a variable, re-running the forest and comparing the decrease in accuracy. Several sources recommend cforests over other machine learning variable importance methods due to its unbiased qualities [@Strobl2007][@Strobl2009][@Hothorn2006].

@@

Limiting the predictive model to a concise set of meaningful variables reduces noise and improves predictions. Less variables means that a simpler model is being used for prediction which is easier to understand for stakeholders [@Weisberg2005]. For these reasons, a subset of the most important variables were chosen before modelling began. This entailed first eliminating outliers, then comparing important variables from ANOVA as well as random forests and cforests. The literature implies that cforests would suit the data for this project as it is unbiased and not limited to linear theory [@Strobl2007].

@@

Previous studies predicting project profitability in software and construction projects were largely limited to case-based reasoning, regression and neural networks. There is a need to test more sophisticated machine learning algorithms that are as powerful as neural networks but provide insight into the reasoning behind predictions. Ensemble trees (boosted trees and random forests) and bayesian networks fit these criteria. Regression and naiive bayes models were also included as simple baseline models. Complex models should be measured against simple models that can be computed at a fraction of the cost. The complete list of models for this study are:

* Bayesian network
* Random forest
* Gradient boosted trees
* Regression - baseline model
* Naiive bayes - baseline model

Some models have limitations and required additional data processing steps such as normalising numeric variables (Bayesian Networks and Regression) and discretising continuous variables (Bayesian Networks). Discretising numeric variables was performed by generating a hierarchical dendrogram of each variable to visualise the clusters. Between four and six clusters were chosen and summarised to find the maximum and minimum values within each cluster. Then, each job was assigned a cluster label to replace a numeric variable. It was decided that for timespans, the discretised variable should be applied to all models because it is easier for a manager to predict a time category than the number of days. For example, a small job could be confidently assigned to less than three weeks and a large job could be assigned to 1.5 - 3 years.

All methods except Gradient Boosted trees could not handle missing values. Preliminary runs of each method used subsets of the data without missing values. Then, imputing methods were trialled with gradient boosted trees since boosted trees can handle missing values. The MICE random forest method has been proven to work well with complex data sets [@Shah2014]. It works by first performing a standard random forest imputation of missing values which are treated as 'place holders'. Then, the imputations for one variable are deleted and the remaining full variables are used to impute the single variable using random forest imputation. This is repeated for each variable and their 'place-holder' imputations are replaced by an imputation targeted at the single variable. This cycle is repeated five times by default [@Azur2011]. If the imputed dataset gave similar predictive results using boosted trees as unimputed data using boosted trees, the imputed data must be reasonable. It is then trialled on the remaining methods and compared to un-imputed trials.

## Model Comparison, Model Averaging

To compare the models, mean squared error statistic was used for regression models and the area under the receiver operating characteristic (ROC) curve (AUC) statistic was used for binary classification. AUC is a better statistic than classification accuracy in comparing algorithms when the probability of the outcome is relevant to the problem. This is true for problems where the decision can depend on whether the probability is above a certain threshold [@Huang2005].

An ROC graph plots the false positive rate on the x-axis and true positive rate on the y-axis where:

@@

After a model has been created using a training set, the test set is run through the model, but in order to classify the test set, a probability threshold must be chosen. That is, the model outputs a probability to classify a case as 0 or 1. The probability is not a true statistical probability, but instead a measure of the  certainty of the model. If a probability threshold is chosen to be 0.6, each case in the test set can then be classified as 0 or 1 depending on whether the probability calculated by the model is greater or less than 0.6. Once the classifications have been made, the true positive rate and false positive rate can be calculated for that model and threshold. This is plotted as a single point on the ROC curve, say point C. Refer below:

@@

A model that is perfectly classified would have all positives correctly classified (1.0 True Postiive Rate) and no incorrectly classified positives (0.0 False Positive Rate). If a model has a 1.0 TPR and 1.0 FPR (top right corner) this means it is correctly classifying all positives at the expense of incorrectly classifying all negatives as positive. It is 'dumbly' classifying all cases as positive. if a model classifies 0.8 of its positives as TP but also 0.8 of its negative cases as positive it is 'dumb' in a similar way to the previosu example. There is an 80% chance any case will classified positive. Therefore, models that lie on the diagonal line classify cases as well as random chance.

To plot the curve in an ROC, points are plotted for results on the training set using numerous threshold probability values between 0 and 1. Once enough points are plotted, a curve can be drawn. The closer the curve reaches the top left hand corner, the closer the algorithm is to perfectly predicting positive and negative cases. This would give an area under the curve (AUC) of 1. AUC of between 0.5 and 1 means the model is performing better than random chance for all possible probability thresholds.

@@

In order to compare whether models perform significantly better than one another, the AUC or mean squared error results from multiple-fold cross validations were used as the data points. Once the best two or three models were determined, these were combined using model averaging. Finally, the best model is determined between the single models and model averaged model.

## Bottom Line Translation

Although predictive power of the final model is important, for the method to be integrated into a company's decisions, the affect on a business' bottom line should be tested. This was done using a profit curve - a chart that plots the amount of profit the company earns on the y axis vs the probability threshold on the x axis.

A simple approach was taken for this analysis, where projects with a probability to be a loss-making job greater than the threshold were rejected entirely. Therefore all profits and losses from jobs above threshold were discounted. If the threshold is zero, all jobs are rejected and the profit is $0. If the threshold is 1.0, all jobs are accepted and the profit is the same as the profit the company actually experienced since the data is a sample of historic projects. The aim was to find the optimal threshold point where saying no to a job above that level will mean result in higher profits, becasue jobs that are likely to make a loss are being rejected. This chart will clarify what percentage of profit increase the company can expect by integrating the algorithm into decision making.

A profit curve is made from a single train/test instance of the data. In order to understand the uncertainty around the profit curve, hundred-fold train/test splits were run and results calculated for each model and threshold point. This resulted in a statistical confidence interval around each threshold point which can be represented graphically. The final profit curve and surrounding confidence interval will indicate whether the algorithm makes a compelling case for integration into the company's decision making process.

>>>>

ANOVA - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.
B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
DAG - directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.
ROC - Reciever Operating Characteristic
AUC - Area under the curve. In this project, the area refers to the area under an ROC curve
TPR - True positive rate
FPR - False positive rate
TP - true positive
FP - false positive
DA - Discriminant analysis
CLV - Customer Lifetime Value
IQR - Interquartile range
ESS - error sum of squares
RMSE - root mean squared error

@@

ANOVA is closely related to linear fit models but incorporates the analysis of differences in group means (i.e. categorical variables) [@Lunney1970]. The method is simple and fast but assumes variables are normally distributed and linearlly related. The variables must therefore be normalised before analysis to achieve a statistical distribution as close to gaussian as possible. A disadvantage of this method is that linear relationships and normally distributed variables do not necessarily represent real world data accurately [@Breiman2001b].

In order to compare variable importance, two types of output from the model are assessed which include the linear coefficients for each variable and the p-values for the coefficients [@Markham2016]. The magnitude of a coefficient does not indicate importance relative to the other variables because its value is directly reflects the values within the variable. The sign of the coefficient however indicates whether the variable is directly or inversely related to the response variable. A p-values for each variable coefficient is output as the result of a significance test for whether the coefficient equals zero. Therefore, if the null hypothesis is rejected (p value < 0.05), it indicates that changes in the explanatory variable are related to changes in the response variable. P-values are a good indication of whether variables have a relationship with the target variable, however they do not necessarily rank importance. If a variable is measured more precisely, it will have a smaller p-value, whereas a variable measured roughly will have a higher p-value. This does not necessarily make the more precisely measure variable is more important than the other. It is also important to keep in mind that, with a dozen variables, chance alone can produce a variable with a p-value < 0.001 7% of the time [@Rice1989].

@@

The random forest algorithm can produce a permutation variable importance for each variable as part of its output. Variable importance is represented as a score, which is calculated by performing permutations on each variable in the test data for each tree and re-running the test data through the trees. More specifically, consider a single variable. The values of this variable in the out-of-bag sample (the test data that the random forest algorithm partitions before building each tree) are randomly permuted (reordered) for each tree. The purpose of permuting is to mimic the absence of that variable. The trees in the random forest are then run again. Variable importance can now be derived by comparing the results of the forest where the covariate has been permuted to the original random forest. If the accuracy of the premuted forest is much less than the original forest, then that variable was important. Permutation importance scores are calculated as the mean decrease in accuracy over all trees for that permuted variable in the random forest [@Breiman2005].

The random forest variable importance provides valuable insight as
it addresses the impact of each predictor variable individually as well as in multivariate interactions with other predictor variables [@Strobl2007]. However, a severe disadvantage of random forest variable importance is that it is not reliable for variable selection where explanatory variables vary in their scale of measurement or their count of categories. Also, the importance of correlated predictors is overestimated and the algorithm tends to favour variables that have many possible splits or many missing values [@Strobl2007]. The intuition behind this is that if a variable has more points to split the data, there are more opportunities for it to split the response variable favourably, however this does not necessarily indicate the variable impacts the response variable most. Overall, the random forest permutation importance is useful as it is able to compare variables in a machine learning environment and can include multivariate interactions, but the current random forest importance measure overestimates the importance of variables with many categories and some numeric variables.

@@

Cforests are an alterantive ensemble tree method to random forests that overcome the aforementioned shortfalls of random forest variable importance measures. It is built from ctrees which are decision trees based on a conditional inference framework. The key difference between ctrees and standard decision trees is that a significance test is used for splitting instead of a purity measure such as the Gini coefficient, which is centered around information gain.

The procedure behind the significance test at each split in a ctree is as follows. To determine which explanatory variable should be used at each split, each variable is permuted in every possible way, and a correlation value is calculated between the tested variable and response variable, for each permutation. The unchanged variable correlation is then compared with the correlation values for all permutations of that variable. From this, a p-value for the true correlation value compared to the permuted correlation values can be calculated. The predictor variable with the lowest p-value is then selected as the splitting variable [@Hothorn2006]. Using an ensemble forest of these trees, variable importance is then calculated in the same fashion as random forests, via permuting a variable, re-running the forest and comparing the decrease in response variable accuracy. Several sources recommend cforests variable importance methods because its statistical p-value tests at each split removes the bias present in other tree ensemble variable importance outcomes such as random forests (as discussed) as well as boosted trees [@Strobl2007][@Strobl2009][@Hothorn2006].

@@

Limiting the predictive model to a concise set of meaningful variables reduces noise and improves predictions. Less variables means that a simpler model is being used for prediction which in turn is easier for stakeholders to understand [@Weisberg2005]. For these reasons, a subset of the most important variables were chosen before modelling which entailed first eliminating outliers, then comparing important variables from ANOVA as well as random forests and cforests. It can be inferred from the literature that cforests would suit the case study data best as it is unbiased and not limited to linear theory [@Strobl2007].

@@

Previous studies predicting project profitability in software and construction projects were predominantly limited to case-based reasoning, regression and neural networks. There is a need to test other sophisticated machine learning algorithms that are as powerful as neural networks but provide insight into the reasoning behind predictions. Ensemble trees (boosted trees and random forests) and bayesian networks fit these criteria. Regression and naiive bayes models were also included as simple baseline models. Complex models should be measured against simple models that can be built at a fraction of the computational cost. The complete list of models tested in this study are:

* Regression - baseline model
* Naiive bayes - baseline model
* Bayesian network
* Random forest
* Gradient boosted trees

Some models have limitations and required additional data processing steps such as normalising numeric variables (Bayesian Networks and Regression) and discretising continuous variables (Bayesian Networks). Discretising numeric variables was performed by generating a hierarchical dendrogram of each variable to visualise the clusters. Between four and six clusters were chosen and summarised to find the maximum and minimum values within each cluster. For example, the diagram below illustrates the hierarchical dendrogram for timespan with 6 clusters boxed.

![](images/method/dendrogram.png)

The number of clusters was determined both visually from the dendrogram and experimentally. Visually, the height of the dendrogram 'branches' indicates the change in 'tightness' of the data points to their cluster centroids as the number of clusters increases. Tightness in the case of Ward's method is the sum of squared distances of each data point to the centroid of its respective cluster (ESS). For example, the height of the top horizontal bar indicates the ESS for one cluster and the height of the second from top horizontal bar is the ESS for two clusters. The hieght of the vertical branch between the two bars indicates the improvement in ESS by adding a cluster [@Putler2012]. As more clusters are added, decreases in ESS are less significant and the aim is to choose the number of clusters at an optimal level in this process.  After initial comparisons of prediction methods had been trialled, the number of clusters for each discretised variable was re-assessed by trialling one more or one less cluster and comparing model output.

Once the number of clusters had been decided, each case was assigned a cluster label to replace its numeric variable. It was decided that for timespans and invoiced amounts, the discretised variables should be applied to all models (not just bayesian networks) because manager's must guess these values initially when operating the model. It is easier for a manager to predict a timespan category than the exact number of days a project will last. For example, a small job could be confidently assigned to less than three weeks and a large job could be assigned to 1.5 - 3 years.

The revised discretised categories for timespan and total amount invoiced were as follows:

| Timespan Categories  | Total Invoiced Amount Categories |
|----------------------|----------------------------------|
| 1 day - 3 weeks      | \$100 - \$600                    |
| 3 weeks - 2.5 months | \$600 - \$2,500                  |
| 2.5 - 9 months       | \$2,500 - \$8,000                |
| 9 months - 1.5 years | \$8,000 - \$60,000               |
| 1.5 - 3 years        | \$60,000 - \$1.8m                |
| more than 3 years    |                                  |

### Missing Data Imputation

All methods except gradient boosted trees could not handle missing data. Therefore, preliminary runs of each method used subsets of the data which had complete data. Then, imputation methods were trialled against gradient boosted tree results since boosted trees can handle missing values and should perform equally well with imputed data.

The MICE random forest method was chosen because it has been proven to work well with complex data sets [@Shah2014]. It first performs a standard random forest imputation of missing values which are treated as initial 'place holders'. Random forest imputation is a complex process which works by first doing a rough imputation of the missing values. Then a ten-tree forest is run to calculate proximities between all cases. Proximities between two cases represent the number of trees in a forest where the two cases were assigned to the same end node in a tree. Once proximities are calculated, missing values for a variable (say variable m) are filled by an average of the other complete m values weighted by the proximity of the case representing each m value.

After the initial random forest imputation is complete, the MICE process continues. The imputations for one variable are deleted and the remaining full variables are used to impute the missing values from the single variable again using random forest imputation. This is repeated for each variable and their 'place-holder' imputations are replaced by an imputation targeted at the single variable. This cycle is repeated five times by default [@Azur2011].

The final imputed dataset was fed into a boosted tree algorithm and if similar predictive results were obtained using boosted trees imputed data and unimputed data, the imputed data must be reasonable. The imputed dataset was then trialled on the remaining methods and compared to un-imputed trials.

## Model Comparison

To compare the models, root mean squared error (RMSE) statistic was used for regression models and the area under the receiver operating characteristic (ROC) curve (AUC) statistic was used to measure binary classification. For binary classification, AUC is a more useful statistic than classification accuracy when output is available as a probability which can be applied to the problem. This is pertinent when it is useful to rank the outcomes and trial probability thresholds that will optimise the rates of true positives and true negatives [@Huang2005]. Note the 'probability' outputs from the models are not true statistical probabilities, but a score the model has assigned to a case using its own measure of certainty.

An ROC graph visualises the curve from which an AUC score is calculated. Its two axes are the false positive rate (FPR) on the x-axis and true positive rate (TPR) on the y-axis. In this problem a 'positive' is a loss-making job so a true positive (TP) is a case when the model predicts a job will be loss making, and it did indeed lose money. A false positive (FP) is when the model predicts a job will be loss making but in reality it was profitable. These values are calculated as follows:

@@

Only predictions made on the test data should be entered into the equations above, therefore after a model must be first created using a training set, then the test data is input. From the test data, the model outputs probability values between 0 and 1 for each case, which cannot be entered into the equations above. In order to classify each case in the test set as 0 or 1, a probability threshold must be chosen. That is, if a probability threshold is chosen to be 0.6, each case in the test set can then be classified as 0 or 1 depending on whether the probability calculated by the model is greater or less than 0.6. Once the classifications have been made, the true positive rate and false positive rate can be calculated for that model and threshold. This is plotted as a single point on the ROC curve, say point C. Refer the diagram below:

@@

A model that is perfectly classified would have all positives correctly classified (1.0 True Postiive Rate) and no incorrectly classified positives (0.0 False Positive Rate). If a model has a 1.0 TPR and 1.0 FPR (top right corner) this means it is correctly classifying all positives at the expense of incorrectly classifying all negatives as positive. It is 'dumbly' classifying all cases as positive. if a model classifies 0.8 of its positives as TP but also 0.8 of its negative cases as positive it is 'dumb' in a similar way to the previous example. There is an 80% chance *any* case will be classified as positive. Therefore, models that lie on the diagonal line classify cases as well as random chance.

To plot the curve in an ROC, a series of points and their coordinates are first calculated. This is done by choosing numerous threshold probability values between 0 and 1, then applying each threshold point to the probability outcomes in the test set, and calculating the TPR and FPR for that threshold point. Once enough points are plotted, a curve can be drawn. The closer the curve reaches toward the top left hand corner, the closer the algorithm is to perfectly predicting positive and negative cases at some optimal threshold. This would give an AUC of 1. An AUC of between 0.5 and 1 means the model is performing better than random chance (as previously stated the diagonal line represents a model randomly assigning positives to all cases at a certain rate).

@@

In order to compare which models performed significantly better than others, an adequate sample size of results statistics was required. Multiple models could be made by using different data in the training vs. testing sets, each providing a resulting test statistic (RMSE or AUC). 5 fold cross validation was used, which created 5 models, then the division of 5 folds was repeated using a different random seed to create another set of 5 training/testing data sets. Initially, 20 models of each method were created in this fashion. Then a two-sample power calculation was run using the two sets of 20 results to determine the sample size to achieve a statistical power of 0.8. Once the number of required samples was determined, more models were built via 5 fold cross validation to obtain additional results statistics. The highest performing models were advanced to the next stage where various methods of combining, or blending, model results were tested.

## Model Blending

- simple averaigng, log reg, FWLS, boosted trees random forests
- repeat 20 times - power - create more samples
- carry all forward for bottom line translation anyway

## Profit Curve

Although predictive power of the final model is important, for the method to be integrated into a company's decisions, the affect on a business' bottom line was tested. This was done using a profit curve - a chart that plots the amount of profit the company earns on the y-axis vs. the probability threshold on the x-axis.

A simple approach was taken for this analysis, where projects with a probability to be a loss-making job greater than the threshold were rejected entirely. Therefore all profits and losses from jobs above the threshold were discounted. If the threshold was zero, all jobs were rejected and the profit would be $0. If the threshold was 1.0, all jobs were accepted and the profit would be the same as the profit the company actually experienced since the data is a sample of historic projects. The aim was to find the optimal threshold point where saying 'no' to a job above that level would result in higher profits, because jobs that were likely to make a loss were being rejected. This chart will clarified what percentage of profit increase the company could expect by integrating the algorithm into decision making. It also provides a clear way to implement the algorithm and promote industry uptake of the research.

Finally, since a profit curve is made from a single training/testing instance of the data, the curve varies with different divisions of the data. Therefore, in order to understand the uncertainty around the profit curve and to determine which blended or individual model performed statistically better than others, a large sample size of curve was required. Again, 20 curves of each method were built and a power calculation was performed to achieve a power of 0.8. This determined the required sample size which was achieved by repeating 5-fold training/testing splits. This enabled the highest points of each curve to be statistically compared to one another. A 95% confidence interval also be determined around the highest point on each curve. The final expected increase in profit and the percentage of projects to be rejected presented a clear scenario that the case study business manager's could assess in terms of their business strategy.
